---
title: 'Part 2: Basic Inferential Data Analysis'
author: "Endri Raco"
date: "4/11/2020"
output: pdf_document
---

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
if(!require(tidyverse)) install.packages('tidyverse', 
repos = 'http://cran.us.r-project.org')
if(!require(gridExtra)) install.packages('gridExtra', 
repos = 'http://cran.us.r-project.org')
if(!require(DataExplorer)) install.packages('DataExplorer', 
repos = 'http://cran.us.r-project.org')
## specify global chunk options
knitr::opts_chunk$set(fig.width = 5, fig.height = 4, dpi = 300,
                      out.width = '90%', fig.align = 'center',
                      tidy.opts=list(width.cutoff=60),
                      tidy=TRUE,
                      cache = TRUE,
                      message = FALSE,
                      warning = FALSE)
```


## Importing data 

```{r, load-data}
# Load ToothGrowth data
library(datasets)
data(ToothGrowth)
```



## Exploratory data analyses

### Overview of the data

Let's understand the dimensions of the dataset, variable names, overall missing summary and data types of each variables

```{r, data-overview}
## overview of the data; 
glimpse(ToothGrowth)
## structure of the data    
introduce(ToothGrowth)
```

Let's show the result of **introduce** by plotting


```{r, data-plot}
plot_intro(ToothGrowth)
```


Now let's plot missing values

```{r, missing-values}
plot_missing(ToothGrowth)
```


## Data Summary

Let's understand data distribution

```{r, data-dist}
# Plot histogram
plot_histogram(ToothGrowth)
```

In similar way let's plot densities

```{r, density}
# Plot densities
plot_density(ToothGrowth)
```

and boxplots:

```{r, boxplot}
plot_boxplot(ToothGrowth, by= 'supp',  ncol = 2)
```

Finally let's understand if variables are correlated:

```{r, corr}
# Correlation plot
plot_correlation(ToothGrowth, cor_args = list( 'use' = 'complete.obs'))
```

## Comparision of tooth growth by supp and dose. 

We will use **t.test** to check if there are group differences due to different supplement type. To perform this test we will assume initially unequal variances between the two groups

```{r, t-test}
# T test
t.test(len ~ supp, data = ToothGrowth)
```

We get a $p-value = 0.06063$, and a confidence interval $(-0.1710156  7.5710156)$ which contains zero. 

This indicates that we can not reject the null hypothesis. It means that the different supplement types have no effect on tooth length.

Next step is to split the data into 3 subsets. Each subset corrensponds to one dosage. 


```{r, data-sub}
# Subset 1
dosage_f <- subset(ToothGrowth, dose %in% c(0.5, 1.0))
# Subset 2
dosage_s <- subset(ToothGrowth, dose %in% c(0.5, 2.0))
# Subset 3
dosage_t<- subset(ToothGrowth, dose %in% c(1.0, 2.0))
```

Let's perform t-test for each of the subsets.

```{r, t-sub1}
# T test for subset 1
t.test(len ~ dose, paired = F, var.equal = F, data = dosage_f)
```

The confidence interval $(-11.983781  -6.276219)$ gives reason the rejection of the null hypothesis. It means there is significant correlation between tooth length and dose levels.

```{r, t-sub2}
# T test for subset 2
t.test(len ~ dose, paired = F, var.equal = F, data = dosage_s)
```
The confidence interval $( -18.15617 -12.83383)$ gives reason the rejection of the null hypothesis. It means there is significant correlation between tooth length and dose levels.

```{r, t-sub3}
# T test for subset 3
t.test(len ~ dose, paired = F, var.equal = F, data = dosage_t)
```

The confidence interval $(-8.996481 -3.733519)$ gives reason the rejection of the null hypothesis. It means there is significant correlation between tooth length and dose levels.


## Assumptions vs Conclusions
    
    - We made the assumptions that populations were indipendent.

    - We reached the conclusion that increase of dosage leads to an increase in tooth growth 
    
    - Supplement has no effect on tooth growth.

